Higher Education Synthesized

### 2025-03-18 artifacts/report/msds_data_mining_project_report.tex
revisions to final report

\begin{abstract}
    American higher education is coming to an impasse. Colleges and universities have seen enrollments grow as larger generations of Americans made their way through postsecondary education. Schools increased their academic offerings and student life activities. Faculty and staff members filled the needs for supporting increasing student enrollments. But there was an issue hiding behind all of this growth: what happens when student enrollments systematically fall?

    The United States has seen declines in high school graduates and, therefore, reduced college applicants. Depending on who you listen to, this ranges from a demographic cliff ushering in an emergency in higher education to a slower decline in higher education enrollments that will play out over years. It is unclear what effect this will have on the near future of higher education. One expected consequence of these declines is that we will see smaller amounts of higher education data relative to previous years.

@@

    The general approach for the work in this project is to recreate a full data pipeline on a smaller scale to demonstrate what we could build out further in a production system. We use Python scripts to simulate pulling data from application programming interfaces (APIs), run databases in Docker --- including a very small scale warehouse --- to stand in for real-world database usage, and leverage Jupyter Notebooks for exploratory data analysis (EDA) and modeling, including our core focus of generative modeling to create synthetic tabular data. We hold off on saving the final tuned versions of models and putting them into automated pipelines, but that would be an extension that we would include in a production system. The main idea is to have all the pieces that we would normally tap into in a regular-enough end-to-end data system to provide a foundation from which to expand on later.

@@

        IPEDS data already comes in a relatively relational format, and that lends itself well to loading it directly into a relational database serving as a data lake. We often work with databases in real-world projects, so it makes sense to include one here. We run a PostgreSQL database locally in Docker. Because Postgres connections are database-specific, we use one shared databse for this layer of the data lake and the coming data warehouse layer, using schemas to separate out the data lake from the data warehouse. This allows us to use structured query language (SQL) scripts to move data between the data lake and data warehouse without needing to use any other tooling, simplifying the handling of data in the database.

@@

        Next, we have two sets of regressions focusing on enrollment and completions. We split each of these two into three different training sets: the real data, the synthetic data, and a combination of the real and synthetic data. This leaves us with six total models to evaluate. We use mean squared error (MSE) and adjusted r-squared ($R^2_a$) to evaluate each model. The focus here is on checking if we see similar values for each of these metrics across all three training sets for a given model. If we do see this, then we have some evidence that the synthetic data is resulting in similar evaluative outputs as the real data, either on its own or in combination with the real data.

@@

    We see expected progression for the generator and discriminator loss. The discriminator loss stays relatively flat while the generator loss decreases. GAN training is notoriously difficulty, including for our use case, so it is promising that we see expected losses in a relatively short time. The final CTGAN we train finishes training in under a minute.

@@

    Next we finish up with machine learning score, training regressors on the real data, synthetic data, and a combination of both and comparing the performance of them with one another. If we find the MSE or $R^2_a$ is similar between the three training sets, then we have some evidence that the synthetic data is similar enough to the real data to stand in for it or to be combined with it without reducing model performance.

    \begin{table}
        \caption{Machine Learning Score}
        \begin{tabular}{llll}
            \toprule
            Metric & Training Set & Target & Value \\
            \midrule
            MSE & Real & Enrollment & 101,200,325.55 \\
            MSE & Synthetic & Enrollment & 111,006,681.53 \\
            MSE & Combined & Enrollment & 101,978,162.27 \\
            $R^2_a$ & Real & Enrollment & 0.19 \\
            $R^2_a$ & Synthetic & Enrollment & 0.11 \\
            $R^2_a$ & Combined & Enrollment & 0.19 \\
            MSE & Real & Completions & 116,066,105.90 \\
            MSE & Synthetic & Completions & 133,327,873.83 \\
            MSE & Combined & Completions & 129,922,536.22 \\
            $R^2_a$ & Real & Completions & -0.01 \\
            $R^2_a$ & Synthetic & Completions & -0.07 \\
            $R^2_a$ & Combined & Completions & -0.04 \\
            \bottomrule
        \end{tabular}
    \end{table}

    The values for each of the machine learning score metrics are pretty similar, hinting that the synthetic data is of pretty good quality. These are not necessarily the best linear regression models in terms of predicting enrollment or completion numbers, but the fact that the MSE and $R^2_a$ values are similar across the real, synthetic, and combined datasets means the generative modeling did its job.

@@

    Finally, a big area of scope reduction is at the end of the pipeline. I originally intended to implement both generative modeling and analytical modeling. Generative modeling is key to the focus of this project, but I was unclear if the project would feel incomplete if it did not have extra layers of analytical modeling as well. It was not until this final update for the project and deliverables that I decided to stick with the generative modeling and leave the analytical modeling until a future expansion of the project. Since I decided to explore generative modeling early in the planning stages, that took priority over analytical modeling, and the generative modeling itself took a significant chunk of time to familiarize myself with, experiment with, and implement. To add extra analytical modeling at this stage feels like it would balloon the scope and pull the focus of the project elsewhere.

@@

    Overall, we have a full pipeline with hooks in it for future improvement, successful implementation of generative modeling along with evaluative metrics that support that claim, the ability to increase the scope of IPEDS data we work with, and a range of other questions and enhancements we can dig into after this iteration of the project. Incorporating generative modeling did add a fresh aspect for me into this work, and I am glad I did it, though it meant that I had to descope other areas to accommodate the new modeling steps.

>>>>

\begin{abstract}
    American higher education is coming to an impasse. Colleges and universities have seen enrollments grow as larger generations of Americans made their way through postsecondary education. Schools increased their academic offerings and student life activities. Faculty and staff members filled the needs for supporting increasing student enrollments. But there was an issue hiding behind all of this growth: what happens when student enrollments systematically begin to drop?

    The United States has seen declines in high school graduates and, therefore, reduced college applicants. Depending on who you listen to, this ranges from a demographic cliff ushering in an emergency in higher education to a slower decline in higher education enrollments that will play out over years. It is not clear what effect this will have on the near future of higher education. One expected consequence of these declines is that we will see smaller amounts of higher education data relative to previous years.

@@

    The general approach for the work in this project is to recreate a full data pipeline on a smaller scale to demonstrate what we could build out further in a production system. We use Python scripts to simulate pulling data from application programming interfaces (APIs), run databases in Docker --- including a very small scale warehouse --- to stand in for real-world database usage, and leverage Jupyter Notebooks for exploratory data analysis (EDA) and modeling, including our core focus of generative modeling to create synthetic tabular data. We hold off on saving the final tuned of models and integrating them into automated pipelines as a future enhancement. The main idea is to have all the pieces that we would normally tap into in a regular-enough end-to-end data system to provide a foundation from which to expand on later.

@@

        IPEDS data already comes in a relatively relational format, and that lends itself well to loading it directly into a relational database serving as a data lake. We often work with databases in real-world projects, so it makes sense to include one here. We run a PostgreSQL database locally in Docker. Because Postgres connections are database-specific, we use one shared database for this layer of the data lake and the coming data warehouse layer, using schemas to separate out the data lake from the data warehouse. This allows us to use structured query language (SQL) scripts to move data between the data lake and data warehouse without needing to use any other tooling, simplifying the handling of data in the database.

@@

        Next, we have two sets of regressions focusing on enrollment and completions. We split each of these two into three different training sets: the real data, the synthetic data, and a combination of the real and synthetic data. This leaves us with six total models to evaluate. We use mean squared error (MSE) and adjusted r-squared ($R^2$) to evaluate each model. The focus here is on checking if we see similar values for each of these metrics across all three training sets for a given model. If we do see this, then we have some evidence that the synthetic data is resulting in similar evaluative outputs as the real data, either on its own or in combination with the real data.

@@

    We see expected progression for the generator and discriminator loss. The discriminator loss stays relatively flat while the generator loss decreases. GAN training is notoriously difficult, including for our use case, so it is promising that we see expected losses in a relatively short time. The final CTGAN we train finishes training in under a minute.

@@

    Next we finish up with machine learning score, training regressors on the real data, synthetic data, and a combination of both and comparing the performance of them with one another. If we find the MSE or $R^2$ is similar between the three training sets, then we have some evidence that the synthetic data is similar enough to the real data to stand in for it or to be combined with it without reducing model performance.

    \begin{table}
        \caption{Machine Learning Score}
        \begin{tabular}{llll}
            \toprule
            Metric & Training Set & Target & Value \\
            \midrule
            MSE & Real & Enrollment & 101,200,325.55 \\
            MSE & Synthetic & Enrollment & 111,006,681.53 \\
            MSE & Combined & Enrollment & 101,978,162.27 \\
            $R^2$ & Real & Enrollment & 0.19 \\
            $R^2$ & Synthetic & Enrollment & 0.11 \\
            $R^2$ & Combined & Enrollment & 0.19 \\
            MSE & Real & Completions & 116,066,105.90 \\
            MSE & Synthetic & Completions & 133,327,873.83 \\
            MSE & Combined & Completions & 129,922,536.22 \\
            $R^2$ & Real & Completions & -0.01 \\
            $R^2$ & Synthetic & Completions & -0.07 \\
            $R^2$ & Combined & Completions & -0.04 \\
            \bottomrule
        \end{tabular}
    \end{table}

    The values for each of the machine learning score metrics are similar, hinting that the synthetic data is of sufficient quality. These are not necessarily the best linear regression models in terms of predicting enrollment or completion numbers, but the fact that the MSE and $R^2$ values are similar across the real, synthetic, and combined datasets means the generative modeling did its job.

@@

    Finally, a big area of scope reduction is at the end of the pipeline. I originally intended to implement both generative modeling and analytical modeling. Generative modeling is key to the focus of this project, but I was unclear if the project would feel incomplete if it did not have extra layers of analytical modeling as well. It was not until this final update for the project and deliverables that I decided to stick with the generative modeling and leave the analytical modeling until a future expansion of the project. Since I decided to explore generative modeling early in the planning stages, that took priority over analytical modeling, and the generative modeling itself took a significant chunk of time to learn, experiment with, and implement. To add extra analytical modeling at this stage feels like it would balloon the scope and pull the focus of the project elsewhere.

@@

    Overall, we have a full pipeline with hooks in it for future improvement. We successfully implemented generative modeling along with evaluative metrics that support that show good performance. We can relatively easily increase the scope of IPEDS data we work with, at least in a technical sense. Incorporating generative modeling did add a fresh aspect for me into this work, and I am glad I did it, though it meant that I had to descope other areas to accommodate the new modeling steps.
